Gaussian Process regression for astronomical time-series

### 2022-11-10 src/tex/ms.tex
[showyourwork] overleaf sync

In other space-based transit surveys such as \textit{CoRoT}, \emph{Kepler} and \emph{TESS}, the systematics are less prominent, so the dominant source of out-of-transit variations is intrinsic stellar variability, which occurs mainly on timescales longer than transits, and can therefore be filtered out quite effectively without removing the transits. However, in some cases, for example for rapidly rotating and magnetically active stars, the separation between transits and stellar signal becomes less clean, significantly reducing the performance of sequential de-trending approaches (irrespective of the algorithm used, \citealt{wotan}).
In such cases, a simultaneous modelling approach could improve detection performance significantly. Since the nuisance signal is specific to each star, it cannot be modelled using other light curves, but a GP is a credible alternative. Such an approach could also offer a small but significant enhancement in the sensitivity future missions such as \textit{PLATO} to transits of habitable planets around Sun-like stars (which are very shallow and relatively long). The main issue would be computational cost: transit surveys monitor $10^4$--$10^5$ stars at a time, with $10^3$--$10^5$ observations per run, and the transit search must be run over a fine grid of periods and phases. Doing this with standard GPR methods would be impractical. However, now fast and scalable GP solvers are available, this becomes a more feasible proposition.

@@

Despite these theoretical advantages and practical successes, the use of GPR to analyse low-resolution transit and eclipse spectra remains confined to a relatively small subset of the corresponding community. Parametric models (in some cases marginalising over families thereof, \citealt{2016ApJ...819...10W}) remain the most widely-used approach for Hubble Space Telescope observations, and Pixel-Level-Decorrelation (PLD) for Spitzer observations \citep{2015ApJ...805..132D}. Possible explanations for this include computational cost, as well as the fact that these other methods are perceived as easier to implement and interpret. Another likely reason, at least in the case of HST transmission spectra, is that GP systematics models tend to produce exoplanet spectra with larger uncertainties than parametric ones. This does not mean that they are less accurate, however, merely that they make a less restrictive set of assumptions about the functional form of the systematics. As discussed by \citet{2014MNRAS.445.3401G} and demonstrated in the appendix of \citet{2018AJ....156..283E}, when the true functional form of the systematics is unknown, choosing the `wrong' systematic model can lead to overconfident estimates of the transit depth, which is more problematic for the robust interpretation of the results than the enlarged uncertainties produced by a more agnostic GP model.
As the recently-launched JWST begins to deliver much higher signal-to-noise observations of exoplanet transits \citep[see e.g.][]{https://doi.org/10.48550/arxiv.2208.11692}, it will be interesting to see how this topic evolves in the next few years. Early results indicates that the systematics are clearly much lower than in e.g.\ HST data, but only time and more data will tell where the systematics floor lies and whether GPR will have a role to play in lowering it.

@@

\subsubsection{Systematics removal in the presence of stellar variability}
GP-based models have also been used to correct instrumental systematics in data from the \textit{K2} space mission, which used the \textit{Kepler} satellite to perform an Ecliptic plane survey after the failure of two of its reaction wheels. During the \textit{K2} observations, the satellite underwent significant roll-angle variations, causing the stars to move on the detector by more than a pixel over a timescale of hours. The satellite thrusters were fired every $\sim 6$\,h to return the spacecraft to its nominal attitude, but the resulting drift caused significant changes in the measured stellar fluxes, due to the detector inter- and intra-pixel variations, and to changes in the contamination of the photometric aperture by neighbouring targets as well as in aperture losses. These can be modelled effectively using a GP with a squared exponential covariance function depending on the roll angle \citep{2015MNRAS.447.2880A, 2016ApJS..226....7C} or the star's 2-D position on the detector \citet{2016MNRAS.459.2408A}. In these approaches, a second, time-dependent term is added to the GP covariance function to represent the target star's intrinsic variability. This not only improves the fit but allows the position-dependent systematics to be evaluated separately and thus removed while preserving intrinsic variability. The use of a time-dependent GP term improves the photometric performance over other widely-used methods for correcting \textit{K2} systematics \citep[e.g.,][]{2014PASP..126..948V} when the stellar variability is significant and/or occurs on timescale similar to the roll angle variations. Ultimately, the best overall photometric precision for \textit{K2} was achieved by combining PLD to model the position-dependent systematics with a time-dependent GP to model intrinsic variability \citep{everest1,everest2}.

@@

\begin{figure}[ht]
  \centering
  \script{scaling.ipynb}
  \includegraphics[width=0.7\linewidth]{figures/scaling.pdf}
  \caption{How the computational cost of a single evaluation of the GP likelihood function (\autoref{eq:gp_like}) scales with the size of the dataset using different algorithms.
  In each case, the kernel function is a one-dimensional Mat\'ern-3/2 with randomly generated data.
  The top two lines labelled ``direct (CPU)'' and ``direct (GPU)'' show the cost of directly evaluating \autoref{eq:gp_like} directly using the Cholesky factorization of $\mathbf{K}$ provided by a tuned and optimised linear algebra library.
  The CPU solver was run using a single-threaded implementation, and the GPU version was executed on an \textsf{A100 NVIDIA} GPU.
  In both cases, the asymptotic scaling of the algorithm is approximately $\mathcal{O}(N^3)$ for $N$ data points.
  The three lines labelled ``celerite ($J$ terms)'' show the performance of the \project{celerite} algorithm \citep[][as implemented by the \project{tinygp} package]{celerite} for a sum of $J$ Mat\'ern-3/2 kernels, since the computational cost of that method scales with the ``complexity'' of the kernel.
  The asymptotic performance of each of these curves is linear in the number of data points, $\mathcal{O}(N)$.}
  \label{fig:scaling}
\end{figure}

>>>>

In other space-based transit surveys such as \textit{CoRoT}, \emph{Kepler} and \emph{TESS}, the systematics are less prominent, so the dominant source of out-of-transit variations is intrinsic stellar variability, which occurs mainly on timescales longer than transits, and can therefore be filtered out quite effectively without removing the transits. However, in some cases, for example for rapidly rotating and magnetically active stars, the separation between transits and stellar signal becomes less clean, significantly reducing the performance of sequential de-trending approaches (irrespective of the algorithm used, \citealt{wotan}).
In such cases, a simultaneous modelling approach could improve detection performance significantly. Since the nuisance signal is specific to each star, it cannot be modelled using other light curves, but a GP is a credible alternative. Such an approach could also offer a small but significant enhancement in the sensitivity of future missions such as \textit{PLATO} to transits of habitable planets around Sun-like stars (which are very shallow and relatively long). The main issue would be computational cost: transit surveys monitor $10^4$--$10^5$ stars at a time, with $10^3$--$10^5$ observations per run, and the transit search must be run over a fine grid of periods and phases. Doing this with standard GPR methods would be impractical. However, now fast and scalable GP solvers are available, this becomes a more feasible proposition.

@@

Despite these theoretical advantages and practical successes, the use of GPR to analyse low-resolution transit and eclipse spectra remains confined to a relatively small subset of the corresponding community. Parametric models (in some cases marginalising over families thereof, \citealt{2016ApJ...819...10W}) remain the most widely-used approach for Hubble Space Telescope observations, and Pixel-Level-Decorrelation (PLD) for Spitzer observations \citep{2015ApJ...805..132D}. Possible explanations for this include computational cost, as well as the fact that these other methods are perceived as easier to implement and interpret. Another likely reason, at least in the case of HST transmission spectra, is that GP systematics models tend to produce exoplanet spectra with larger uncertainties than parametric ones. This does not mean that they are less accurate, however, merely that they make a less restrictive set of assumptions about the functional form of the systematics. As discussed by \citet{2014MNRAS.445.3401G} and demonstrated in the appendix of \citet{2018AJ....156..283E}, when the true functional form of the systematics is unknown, choosing the `wrong' systematic model can lead to overconfident estimates of the transit depth, which is more problematic for the robust interpretation of the results than the enlarged uncertainties produced by a more agnostic GP model.

@@

\subsubsection{Systematics removal in the presence of stellar variability}
GP-based models have also been used to correct instrumental systematics in data from the \textit{K2} space mission, which used the \textit{Kepler} satellite to perform an Ecliptic plane survey after the failure of two of its reaction wheels. During the \textit{K2} observations, the satellite underwent significant roll-angle variations, causing the stars to move on the detector by more than a pixel over a timescale of hours. The satellite thrusters were fired every $\sim 6$\,h to return the spacecraft to its nominal attitude, but the resulting drift caused significant changes in the measured stellar fluxes, due to the detector inter- and intra-pixel variations, and to changes in the contamination of the photometric aperture by neighbouring targets as well as in aperture losses. These can be modelled effectively using a GP with a squared exponential covariance function depending on the roll angle \citep{2015MNRAS.447.2880A, 2016ApJS..226....7C} or the star's 2-D position on the detector \citep{2016MNRAS.459.2408A}. In these approaches, a second, time-dependent term is added to the GP covariance function to represent the target star's intrinsic variability. This not only improves the fit but allows the position-dependent systematics to be evaluated separately and thus removed while preserving intrinsic variability. The use of a time-dependent GP term improves the photometric performance over other widely-used methods for correcting \textit{K2} systematics \citep[e.g.,][]{2014PASP..126..948V} when the stellar variability is significant and/or occurs on timescale similar to the roll angle variations. Ultimately, the best overall photometric precision for \textit{K2} was achieved by combining PLD to model the position-dependent systematics with a time-dependent GP to model intrinsic variability \citep{everest1,everest2}.

@@

\begin{figure}[ht]
  \centering
  \script{scaling.ipynb}
  \includegraphics[width=0.7\linewidth]{figures/scaling.pdf}
  \caption{Computational cost of a single evaluation of the GP likelihood function (\autoref{eq:gp_like}) versus the size of the dataset using different algorithms.
  In each case, the kernel function is a one-dimensional Mat\'ern-3/2 with randomly generated data.
  The top two lines labelled ``direct (CPU)'' and ``direct (GPU)'' show the cost of evaluating \autoref{eq:gp_like} directly using the Cholesky factorization of $\mathbf{K}$ provided by a tuned and optimised linear algebra library.
  The CPU solver was run using a single-threaded implementation, and the GPU version was executed on an \textsf{A100 NVIDIA} GPU.
  In both cases, the asymptotic scaling of the algorithm is approximately $\mathcal{O}(N^3)$ for $N$ data points.
  The three lines labelled ``celerite ($J$ terms)'' show the performance of the \project{celerite} algorithm \citep[][as implemented by the \project{tinygp} package]{celerite} for a sum of $J$ Mat\'ern-3/2 kernels, since the computational cost of that method scales with the ``complexity'' of the kernel.
  The asymptotic performance of each of these curves is linear in the number of data points, $\mathcal{O}(N)$.}
  \label{fig:scaling}
\end{figure}
